---
title: "P8105 Data Science I Homework 3"
author: Olivia Wang (hw2852)
output: github_document
date: "2022-10-15"
---

In preparation for the problems below, we will load the following libraries: 

```{r load_libraries}
library(tidyverse)
```

# Problem 1

This problem uses the `Instacart` data. This data set will be loaded through the `p8105.datasets` library. 

```{r}
library(p8105.datasets)
data("instacart")
```

## Description of `Instacart` Data Set

## Aisles

We can enumerate the number of aisles by applying the `group_by` function to identify the number of unique `aisle_id` variable values. The number of rows generated in the output would be the number of aisles in the data set. Building upon the results generated from the `group_by` function, we may determine the aisles from which the most items were ordered. This process involves generating a summary of the number of times each `aisle_id` appears in these data, then arranging the aisles in decreasing order of the number of times it appears. 

```{r}
instacart %>%
  group_by(aisle_id) %>% 
  summarize(count = n()) %>% 
  arrange(desc(count))
```

There are __134 aisles__ in the `Instacart` data set. Of the 134 aisles, the following are the aisles from which the most items are ordered:

Aisle Number   Number of Items Sold
-------------  --------------------
__83__         150,609
__24__         150,473
__123__        78,493
__120__        55,240
__21__         41,699
